On the entropy of isothermal dusty gas: application to protoplanetary disks

### 2016-08-01 paper.tex
updated notes
order of mag estimate
demo fig
general edits
sample results
notes and code update
title change
text and code update
adding results
regular update

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\rhod}{\rho_\mathrm{d}}
\newcommand{\rhog}{\rho_\mathrm{g}}
\newcommand{\tstop}{t_\mathrm{s}}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\defcitealias{goldreich67}{GS67}

@@

\title{untitled}
%\title{Vertical shear instability in astrophysical disks: vertically
%global linear calculations with thermal relaxation and applications
%to the Solar Nebula}  
\author{Min-Kai Lin\altaffilmark{1}}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
dusty vsi, or the lack thereof
\end{abstract}

\input{model}
%\input{linear_iso}
%\input{linear_adia}
%\input{numerical}
%\input{application} 
%\input{summary}

%\appendix
%\input{appendix}

>>>>

\usepackage{amsmath}
\usepackage{paralist}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\Htilde}{H_\epsilon}
\newcommand{\Hdust}{H_\mathrm{d}}
\newcommand{\Hgas}{H_\mathrm{g}}
\newcommand{\hgas}{h_\mathrm{g}}
\newcommand{\tepsilon}{\tilde{\epsilon}}
\newcommand{\rhod}{\rho_\mathrm{d}}
\newcommand{\rhog}{\rho_\mathrm{g}}
\newcommand{\tstop}{t_\mathrm{s}}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\Hg}{H_\mathrm{g}}
\newcommand{\Hd}{H_\mathrm{d}} 
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\defcitealias{lin15}{LY15}
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\defcitealias{goldreich67}{GS67}

@@

\title{On the entropy of 
  isothermal dusty gas: application to protoplanetary disks}  
\author{Min-Kai Lin\altaffilmark{1}}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
  Small dust particles embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction, causing them to almost follow the gas
  flow. This is similar to the conservation of entropy following an
  adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty-gas indeed behaves like an  
  adiabatic pure gas, while finite dust-gas coupling may be
  interpreted as an effective 
  heating/cooling. We exploit this equivalence 
  to 
\begin{inparaenum}[1)]
\item
  define the entropy of isothermal dusty gas; and 
\item 
  generalize previous, well-known gaseous instabilities in
  protoplanetary disks to include dust dynamics. 
\end{inparaenum}
This includes the axisymmetric stability of strictly isothermal dusty
gas, from which we place quantitative constraints on the sharpness of 
dust edges that can persist in protoplanetary disks. We also extend
the vertical shear instability (VSI) in locally isothermal disks to include
dust. We find that  dust-loading reduces VSI growth rates. The presence of
strongly-coupled dust particles preferentially stabilize VSI modes with small radial
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. We briefly discuss other applications of the 
equivalence between isothermal dusty gas and adiabatic hydrodynamics
to understand other dusty instabilities in protoplanetary disks. 
\end{abstract}

\section{Introduction}

\input{model}
\input{linear}
\input{result}
%\input{linear_adia}
%\input{numerical}
%\input{application} 
%\input{summary}

\appendix
\input{appendix}

### 2016-08-01 result.tex
regular update

\section{Perfectly coupled
  dusty vertical shear instability}\label{results}

When $\tstop=0$ we may construct exact dusty equilibra by specifying
the dust-to-gas ratio, as described in \S\ref{eqm}. We use this to
examine the effect of dust on classic VSI, defined to be driven by the
temperature gradient (\S\ref{dusty_vsi_int}). For simplicity we
consider constant input values of the midplane dust-to-gas ratio
$\tepsilon_0$ and the characteristic dust thickness $\Hd$. These,
together with the perturbation wavenumber $k$, comprises main the
parameters of the linear problem. We fix the radial power-law index
for the midplane gas density to $p = -1.5$, that for the
temperature profile to $q=-1$; and set the gas disk aspect-ratio
$h_\mathrm{g}=0.05$. These are fiducial values used in \citetalias{lin15}.

\subsection{Effect of dust-loading}
We first vary the dust-to-gas ratio $\tepsilon_0\in[10^{-3},1]$,
fixing the dust thickness to $\Hd=0.99\Hg$ (so that $\tepsilon$ is
roughly constant with height) and perturbation
wavenumber to $k\Hg = 30$.

Fig. \ref{vsi_dust_loading} show unstable
modes found for different values of $\tepsilon_0$. The eigenvalue
distributions for $\tepsilon_0 \leq 10^{-2}$ are similar to the
dust-free fiducial case considered by \citetalias{lin15}, consisting
of the roughly horizontal `body modes' and the nearly-vertical
`surface modes'. The latter is due to the imposed vertical boundaries
\citep{barker15}.

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably `surface modes', which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\tepsilon_0\geq
0.1$. The body modes' growth rates remain $\sim O(h\OmK)$ but their
oscillation frequency increases with $\tepsilon_0$.

In Fig. \ref{vsi_dust_loading2d} we compare the

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_kx30Hd1} 
  \caption{Unstable modes in a locally isothermal, perfectly coupled
    dusty disk with fiducial parameters
    $(p,q,h_\mathrm{g}, \Hd/\Hg )=(-1.5,-1,0.05, 0.99)$. The real
    frequency $\omega$ and growth rates $s$ are shown for a range of
    midplane dust-to-gas ratios $\tepsilon_0=\rhod/\rhog$. 
    \label{vsi_dust_loading}
    }
\end{figure}

\begin{figure}
  \includegraphics[scale=0.54, clip=true, trim=0cm 2.5cm 0cm 0cm]{figures/result2d_dg1d-3.ps}\\
  \includegraphics[scale=0.54]{figures/result2d_dg1.ps} 
  \caption{Fundamental dusty VSI mode in real space for midplane dust-to-gas
    ratio $\tepsilon_0=10^{-3}$ (top) and $\tepsilon_0=1$
    (bottom). The color scale shows the perturbation to the
    dust-to-gas ratio, $\delta\tepsilon$; and the arrows show
    $\sqrt{\rho}\left(\dd v_x, \dd v_z\right)$. 
    \label{vsi_dust_loading2d}
    }
\end{figure}

@@

>>>>

\section{
  Dusty vertical shear instability}\label{results}

When $\tstop=0$ we may construct exact dusty equilibra by specifying
the dust-to-gas ratio, as described in \S\ref{eqm}. We use this to
examine the effect of dust on classic VSI, defined to be driven by the
temperature gradient (\S\ref{dusty_vsi_int}). For simplicity we
consider constant input values of the midplane dust-to-gas ratio
$\epsilon_0$ and the characteristic dust thickness $\Hd$. These,
together with the perturbation wavenumber $k$, comprises main the
parameters of the linear problem. We fix the radial power-law index
for the midplane gas density to $p = -1.5$, that for the
temperature profile to $q=-1$; and set the gas disk aspect-ratio
$h_\mathrm{g}=0.05$. These are fiducial values used in \citetalias{lin15}.

\subsection{Effect of dust-loading}
We first vary the dust-to-gas ratio $\epsilon_0\in[10^{-3},1]$,
fixing the dust thickness to $\Hd=0.99\Hg$ (so that $\epsilon$ is
roughly constant with height) and perturbation
wavenumber to $k\Hg = 30$.

Fig. \ref{vsi_dust_loading} show unstable
modes for different values of $\epsilon_0$. The eigenvalue
distributions for $\epsilon_0 \leq 10^{-2}$ are similar to the
dust-free fiducial case considered by \citetalias{lin15}, consisting
of the roughly horizontal `body modes' and the nearly-vertical
`surface modes'. The latter is due to the imposed vertical boundaries
\citep{barker15}.

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably `surface modes', which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\epsilon_0\geq
0.1$. The body modes' growth rates remain $\sim O(h_\mathrm{g}\OmK)$ but their
oscillation frequency increases with $\epsilon_0$. The lowest
frequency `fundamental' body mode is not the fastest growing, but they
are energetically dominant because the entire disk column is perturbed
\citep[cf. surface modes which only disturb the disk boundaries,][]{umurhan16c}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_kx30Hd1} 
  \caption{Unstable modes in a locally isothermal, perfectly coupled
    dusty disk with fiducial parameters
    $(p,q,h_\mathrm{g}, \Hd/\Hg )=(-1.5,-1,0.05, 0.99)$. The real
    frequency $\omega$ and growth rates $s$ are shown for a range of
    midplane dust-to-gas ratios $\epsilon_0=\rhod/\rhog$. 
    \label{vsi_dust_loading}
    }
\end{figure}

In Fig. \ref{vsi_dust_loading2d}
we compare the fundamental mode between the nearly 
dust-free case $\epsilon_0=10^{-3}$ an a dusty disk with
$\epsilon_0=1$. Dust-loading preferentially
stabilizes the disk atmosphere against the VSI, restricting
meridional motions to $|z|\lesssim 2\Hg$, where most of the disk mass
reside. Notice also the perturbed dust-to-gas ratio,
$\delta\epsilon$, shifts from being maximized at the disk boundaries
to the midplane.

\begin{figure}
  \includegraphics[scale=0.54, clip=true, trim=0cm 2.5cm 0cm 0cm]{figures/result2d_dg1d-3.ps}\\
  \includegraphics[scale=0.54]{figures/result2d_dg1.ps} 
  \caption{Fundamental dusty VSI mode in real space for midplane dust-to-gas
    ratio $\epsilon_0=10^{-3}$ (top) and $\epsilon_0=1$
    (bottom). The color scale shows the perturbation to the
    dust-to-gas ratio, $\delta\epsilon$; and the arrows show
    $\sqrt{\rho}\left(\dd v_x, \dd v_z\right)$. 
    \label{vsi_dust_loading2d}
    }
\end{figure}

In Fig. \ref{vsi_dust_loading_vareps} we plot the growth rates as a
function of $\epsilon_0$ for different perturbation wavenumbers $k$. We
find that dust-loading stabilizes the VSI more effectively for shorter
wavelength perturbations. This is because for high wavenumbers the
dominant modes are surface modes, which are stabilized by
dust-loading. The figure suggest that VSI becomes much less efficient
for $\epsilon_0\gtrsim 0.1$ and $k\Hg\gtrsim 50$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_vareps} 
  \caption{Maximum growth rate of the dusty VSI as a function of the
    midplane dust-to-gas ratio $\epsilon_0$ for perturbations with
    different radial wavenumbers $k$. The dust layer thickness is
    fixed to $\Hd\simeq \Hg$. 
    \label{vsi_dust_loading_vareps}
    }
\end{figure}

@@

\subsection{Fixed metalicity}
